---
title: "species categories method"
author: "LU FENG"
date: "2026-05-06"
output: html_document
---
# Main purpose:
# This script identifies bird species whose native extant ranges occur only on islands and continent
# by overlaying BirdLife range maps with global island and mainland polygons.

# Input data:
# 1. BirdLife BOTW_2025.gpkg range maps
# 2. USGS Global Islands geodatabase, including islands and mainlands

# Output:
# species-level island and mainland overlap areas;
# island-only, mainland-only, and island-mainland shared classification;
# diagnostic table for species whose area calculation failed.

```{r package}
# 1. Load required packages and set spatial processing options

library(sf)
library(dplyr)
library(data.table)
library(lwgeom)

# Skip GDAL polygon organization to speed up reading complex polygon data
Sys.setenv(OGR_ORGANIZE_POLYGONS = "SKIP")

# Use planar geometry because all area calculations are done in an equal-area CRS.
sf_use_s2(FALSE)

# Print warnings immediately
options(warn = 1)

```


```{r paths-and-parameters}
project_dir <- path.expand("~/island_bird_project")
# Main project directory

bird_path  <- file.path(project_dir, "BOTW_2025.gpkg")
island_gdb <- file.path(project_dir, "GlbIslands.gdb")
out_dir    <- file.path(project_dir, "output")
chunk_out_dir <- file.path(out_dir, "chunks")
# Output folder for each SLURM array chunk

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(chunk_out_dir, showWarnings = FALSE, recursive = TRUE)
# Create output folders if they do not exist

# Equal-area projection used for area calculation
# EPSG:6933 = World Cylindrical Equal Area
equal_area_crs <- 6933
```

```{r BirdLife filters and SLURM settings, include=FALSE}
# Keep only extant ranges
presence_keep <- c(1)
# keep only native ranges
origin_keep   <- c(1)
# keep only breeding and resident ranges
seasonal_keep <- c(1, 2)

# Mainland overlap tolerance in km2.
# Species with mainland overlap <= this threshold are still treated as island-only.
mainland_tol_km2 <- 1

# Divide the analysis into different tasks to speed up the operation
# Current SLURM array task ID
array_id <- as.integer(Sys.getenv("SLURM_ARRAY_TASK_ID", "1"))
# Number of species processed by each array job
species_per_job <- as.integer(Sys.getenv("SPECIES_PER_JOB", "100"))
# Number of species processed in each spatial chunk
chunk_size <- as.integer(Sys.getenv("CHUNK_SIZE", "10"))
```

```{r check paramters valid, include=FALSE}

# Calculate the SQL offset for this array job
# This determines which subset of species will be processed
offset_n <- (array_id - 1) * species_per_job
job_label <- sprintf("job%03d", array_id)

# Helper function to print timestamped log messages
# This is useful for monitoring long-running jobs on the cluster
log_msg <- function(...) {
  message(format(Sys.time(), "%Y-%m-%d %H:%M:%S"), " | ", ...)
}

# SQL filter for BirdLife range polygons
bird_where <- paste0(
  "presence IN (", paste(presence_keep, collapse = ","), ") ",
  "AND origin IN (", paste(origin_keep, collapse = ","), ") ",
  "AND seasonal IN (", paste(seasonal_keep, collapse = ","), ")"
)

#Inner query: determines which 100 species are assigned to the current array job
#Outer query: reads all BirdLife range polygons for those 100 selected species
bird_query <- paste0("
  SELECT sisid, sci_name, presence, origin, seasonal, geom
  FROM all_species
  WHERE ", bird_where, "
    AND sisid IN (
      SELECT sisid FROM (
        SELECT DISTINCT sisid
        FROM all_species
        WHERE ", bird_where, "
        ORDER BY sisid
        LIMIT ", species_per_job, "
        OFFSET ", offset_n, "
      )
    )
")

log_msg("Reading BirdLife data...")
```

```{r read and clean birdlife data, include=FALSE}

# Read BirdLife polygons for this array job
birds_features <- st_read(
  dsn = bird_path,
  query = bird_query,
  quiet = FALSE
)

# This species was skipped because it takes extremely long to process.
skip_sisid <- c(22680405)

birds_features <- birds_features %>%
  filter(!(sisid %in% skip_sisid))

# If no species are assigned to this job, stop the script
if (nrow(birds_features) == 0) {
  log_msg("No BirdLife features assigned to this array job. Exiting.")
  quit(save = "no", status = 0)
}

# Remove empty geometries
birds_features <- birds_features[!st_is_empty(birds_features), ]

# Repair invalid geometries and transform to equal-area projection
birds_features <- birds_features %>%
  st_make_valid() %>%
  st_transform(equal_area_crs)
```



```{r Dissolve polygons by species, include=FALSE}

# Read BirdLife polygons for this array job
log_msg("Dissolving bird ranges by species...")

#BirdLife may contain multiple polygons for the same species.
#This function combines all polygons belonging to one species into a single geometry.

dissolve_species <- function(x) {
  #Extract species ID and scientific name
  sid <- x$sisid[1]
  sname <- x$sci_name[1]
  
  # Remove empty geometries
  x <- x[!st_is_empty(x), ]
  
  # If all geometries of this species are empty, skip this species
  if (nrow(x) == 0) return(NULL)
  
  # Repair geometries before union
  x <- st_make_valid(x)
  
  # First attempt: dissolve polygons using st_union
  g <- tryCatch(
    st_union(st_geometry(x)),
    error = function(e) NULL
  )
  
  # Fallback: buffer(0) sometimes repairs topology problems
  if (is.null(g)) {
    x2 <- tryCatch(st_buffer(x, 0), error = function(e) NULL)
    if (!is.null(x2)) {
      g <- tryCatch(
        st_union(st_geometry(st_make_valid(x2))),
        error = function(e) NULL
      )
    }
  }
  
  # Fallback 2: if union fails, use st_combine
  if (is.null(g)) {
        g <- tryCatch(st_combine(st_geometry(x)), error = function(e) NULL)
        }
  if (is.null(g)) return(NULL)
  g <- st_make_valid(g)
  
  # Area calculations require polygon geometries.
  if (!inherits(g, "sfc_POLYGON") && !inherits(g, "sfc_MULTIPOLYGON")) {
    g <- st_collection_extract(g, "POLYGON")
  }
  
  #Remove empty geometries after extraction or validation.
  g <- g[!st_is_empty(g)]
  if (length(g) == 0) return(NULL)
  
# If the result contains multiple geometry features, combine them so that
# each species is represented by a single geometry object.
  
if (length(g) > 1) {
    g <- st_make_valid(st_combine(g))
  }
  st_sf(
    sisid = sid,
    sci_name = sname,
    geometry = g,
    crs = st_crs(x)
  )
}

# Split BirdLife range polygons by species ID (sisid)..

birds_species <- split(birds_features, birds_features$sisid) %>%
  lapply(dissolve_species) %>%
  Filter(Negate(is.null), .) %>%
  do.call(rbind, .) %>%
  st_make_valid()

# Total range size is calculated after dissolving each species.
birds_species$total_range_area_km2 <- as.numeric(st_area(birds_species)) / 1e6
log_msg("Species after dissolve: ", nrow(birds_species))

# Remove large intermediate objects to free memory.
rm(birds_features)
gc()

```

```{r Read island and mainland layers, include=FALSE}

# Read only geometry because attribute columns are not needed here.
read_geom_only <- function(gdb, layer_name) {
  x <- st_read(gdb, layer = layer_name, quiet = FALSE)
  st_sf(geometry = st_geometry(x), crs = st_crs(x))
}

# Read island layers of different size classes
big_islands <- read_geom_only(island_gdb, "BigIslands")
small_islands <- read_geom_only(island_gdb, "SmallIslands")
very_small_islands <- read_geom_only(island_gdb, "VerySmallIslands")
mainlands <- read_geom_only(island_gdb, "Mainlands")

# Combine all island layers into one island dataset
islands <- bind_rows(big_islands, small_islands, very_small_islands)

# Remove empty island geometries and transform to equal-area projection
islands <- islands[!st_is_empty(islands), ]

islands <- islands %>%
   st_make_valid() %>%
  st_transform(equal_area_crs)

mainlands <- mainlands[!st_is_empty(mainlands), ]

mainlands <- mainlands %>%
  st_make_valid() %>%
  st_transform(equal_area_crs)

rm(big_islands, small_islands, very_small_islands)
gc()

log_msg("Island polygons: ", nrow(islands))
log_msg("Mainland polygons: ", nrow(mainlands))
```

```{r Checking mainland intersections, include=FALSE}
# This function checks whether each species range intersects a spatial mask.
# The mask can be islands or mainlands.
# The operation is performed in chunks to reduce memory use.

intersection_flag <- function(species_sf, mask_sf, flag_name) {
  # Split species into chunks
  idx <- split(
    seq_len(nrow(species_sf)),
    ceiling(seq_len(nrow(species_sf)) / chunk_size)
  )
  
  # For each chunk, check spatial intersection
  out <- lapply(seq_along(idx), function(i) {
    log_msg("Checking ", flag_name, " chunk ", i, " / ", length(idx))
    
    x <- species_sf[idx[[i]], ]
    
    data.frame(
      sisid = x$sisid,
      
      # TRUE if the species intersects at least one polygon in the mask
      flag = lengths(st_intersects(x, mask_sf)) > 0
    )
  }) %>%
    bind_rows()
  # Rename output column
  names(out)[names(out) == "flag"] <- flag_name
  out
}

log_msg("Checking island intersections...")

# Check whether species ranges overlap islands
island_flag <- intersection_flag(birds_species, islands, "island_overlap")
# Check whether species ranges overlap mainlands
mainland_flag <- intersection_flag(birds_species, mainlands, "mainland_overlap")
```

```{r Calculate mainland overlap area}
# Calculate mainland overlap area for mainland-touching species

# This function calculates the total area of overlap between each species range
  # and a given mask layer.
  #
  # species_sf  = dissolved species range polygons
  # mask_sf     = island polygons or mainland polygons
  # overlap_ids = species IDs that have at least one spatial intersection
  # area_name   = output area column name, e.g. "island_area_km2"
  # status_name = output status column name, e.g. "island_area_status"

calc_overlap_area <- function(species_sf, mask_sf, overlap_ids, area_name, status_name) {
  base <- species_sf %>%
    st_drop_geometry() %>%
    distinct(sisid)
  xall <- species_sf %>%
    filter(sisid %in% overlap_ids)
  # This does not mean there are no input species.
  # It means none of the species in this array job overlaps the current mask
  # layer, such as islands or mainlands. In that case, all species get zero
  # overlap area for this mask.
  if (nrow(xall) == 0) {
    out <- base %>%
      mutate(
        area_km2 = 0,
        area_status = "absent"
      )
    names(out)[names(out) == "area_km2"] <- area_name
    names(out)[names(out) == "area_status"] <- status_name
    return(out)
  }
  idx <- split(
    seq_len(nrow(xall)),
    ceiling(seq_len(nrow(xall)) / chunk_size)
  )
  area_hit <- lapply(seq_along(idx), function(i) {
    log_msg("Calculating ", area_name, " chunk ", i, " / ", length(idx))
    x <- xall[idx[[i]], ]
    # Keep only mask polygons that intersect this species chunk.
    mask_ids <- unique(unlist(st_intersects(x, mask_sf)))
    if (length(mask_ids) == 0) {
      return(data.frame(
        sisid = x$sisid,
        area_km2 = 0,
        area_status = "absent"
      ))
    }
    # Calculate the actual geometric overlap.
    inter <- tryCatch(
      st_intersection(st_make_valid(x), mask_sf[mask_ids, ]),
      error = function(e) NULL
    )
    # If the area calculation fails, keep it as NA instead of treating it as zero.
    if (is.null(inter)) {
      return(data.frame(
        sisid = x$sisid,
        area_km2 = NA_real_,
        area_status = "area_failed"
      ))
    }
    # Keep only polygon overlaps.
    # Boundary-only contacts may produce points or lines, which should not count as area.
    inter <- st_collection_extract(inter, "POLYGON")
    inter <- inter[!st_is_empty(inter), ]
    if (nrow(inter) == 0) {
      return(data.frame(
        sisid = x$sisid,
        area_km2 = 0,
        area_status = "absent"
      ))
    }
    inter %>%
      mutate(area_km2 = as.numeric(st_area(.)) / 1e6) %>%
      st_drop_geometry() %>%
      group_by(sisid) %>%
      summarise(
        area_km2 = sum(area_km2, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(area_status = "present")
  }) %>%
    bind_rows() %>%
    group_by(sisid) %>%
    summarise(
      area_km2 = ifelse(any(is.na(area_km2)), NA_real_, sum(area_km2, na.rm = TRUE)),
      area_status = case_when(
        any(is.na(area_km2)) ~ "area_failed",
        sum(area_km2, na.rm = TRUE) > 0 ~ "present",
        TRUE ~ "absent"
      ),
      .groups = "drop"
    )
  # Add back species with no overlap.
  out <- base %>%
    left_join(area_hit, by = "sisid") %>%
    mutate(
          area_km2 = ifelse(is.na(area_km2) & is.na(area_status), 0, area_km2),
          area_status = ifelse(is.na(area_status), "absent", area_status)
          )
  names(out)[names(out) == "area_km2"] <- area_name
  names(out)[names(out) == "area_status"] <- status_name
  out
}

# Run the mainland&islands area calculation for species that intersect mainland and islands polygons.
log_msg("Calculating island overlap area...")

island_area <- calc_overlap_area(
  species_sf = birds_species,
  mask_sf = islands,
  overlap_ids = island_flag %>% filter(island_overlap) %>% pull(sisid),
  area_name = "island_area_km2",
  status_name = "island_area_status"
)
rm(islands)
gc()

log_msg("Calculating mainland overlap area...")
mainland_area <- calc_overlap_area(
  species_sf = birds_species,
  mask_sf = mainlands,
  overlap_ids = mainland_flag %>% filter(mainland_overlap) %>% pull(sisid),
  area_name = "mainland_area_km2",
  status_name = "mainland_area_status"
)
rm(mainlands)
gc()
```

```{r Classify species as island-only or non-island-only}
log_msg("Classifying species...")

species_result <- birds_species %>%
  st_drop_geometry() %>%
  select(sisid, sci_name, total_range_area_km2) %>%
  left_join(island_flag, by = "sisid") %>%
  left_join(mainland_flag, by = "sisid") %>%
  left_join(island_area, by = "sisid") %>%
  left_join(mainland_area, by = "sisid") %>%
  mutate(
     # If no overlap information is available, treat as FALSE
    island_overlap = ifelse(is.na(island_overlap), FALSE, island_overlap),
    mainland_overlap = ifelse(is.na(mainland_overlap), FALSE, mainland_overlap),
    
    island_area_km2 = ifelse(is.na(island_area_km2) & !island_overlap, 0, island_area_km2),
    mainland_area_km2 = ifelse(is.na(mainland_area_km2) & !mainland_overlap, 0, mainland_area_km2),

    island_area_prop = island_area_km2 / total_range_area_km2,
    mainland_area_prop = mainland_area_km2 / total_range_area_km2,

    island_area_prop = ifelse(is.infinite(island_area_prop), NA_real_, island_area_prop),
    mainland_area_prop = ifelse(is.infinite(mainland_area_prop), NA_real_, mainland_area_prop),

    mainland_area_status_for_classification = case_when(
      is.na(mainland_area_km2) ~ "area_failed",
      mainland_area_km2 <= mainland_tol_km2 ~ "within_tolerance",
      TRUE ~ "above_tolerance"
    ),

    island_only_binary_strict = island_overlap & !mainland_overlap,
    island_only_area_based = island_overlap &
      mainland_area_status_for_classification == "within_tolerance",
    island_only = island_only_area_based
  ) %>%
  arrange(desc(island_only), mainland_area_status_for_classification, mainland_area_km2, sci_name)

island_only_species <- species_result %>%
  filter(island_only)

area_failed_species <- species_result %>%
  filter(
    island_area_status == "area_failed" |
      mainland_area_status == "area_failed" |
      mainland_area_status_for_classification == "area_failed"
  )
```

```{r Write output tables, range maps, and parameter summaries}
fwrite(
  species_result,
  file.path(chunk_out_dir, paste0("all_species_same_method_", job_label, ".csv"))
)

fwrite(
  island_only_species,
  file.path(chunk_out_dir, paste0("island_only_species_same_method_", job_label, ".csv"))
)

fwrite(
  area_failed_species,
  file.path(chunk_out_dir, paste0("area_failed_species_same_method_", job_label, ".csv"))
)

parameter_summary <- data.frame(
  parameter = c(
    "presence_keep",
    "origin_keep",
    "seasonal_keep",
    "mainland_tol_km2",
    "equal_area_crs",
    "array_id",
    "species_per_job",
    "offset_n",
    "chunk_size",
    "n_species_analyzed",
    "n_island_only_candidates",
    "n_area_failed",
    "classification_rule"
  ),
  value = c(
    paste(presence_keep, collapse = ","),
    paste(origin_keep, collapse = ","),
    paste(seasonal_keep, collapse = ","),
    mainland_tol_km2,
    equal_area_crs,
    array_id,
    species_per_job,
    offset_n,
    chunk_size,
    nrow(species_result),
    nrow(island_only_species),
    nrow(area_failed_species),
    "island_overlap == TRUE and mainland_area_status_for_classification == within_tolerance"
  )
)

fwrite(
  parameter_summary,
  file.path(chunk_out_dir, paste0("analysis_parameters_same_method_", job_label, ".csv"))
)

log_msg("Analysis finished.")
log_msg("Output folder: ", chunk_out_dir)
```
